---
title: "Ultimate"
author: "Sander van Beek"
date: "16-5-2020"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###############
# Intialising #
###############

```{r setup, include = FALSE}
require("ggplot2")
```

```{r}
default_states <- 25            # The number of states in the model
default_steps <- 5              # The number of states one descends after a claim
default_prob  <- 0.05           # Average claim frequency
default_gamma <- 2              # Parameter used for the gamma distribution
default_infinitesimal <- 1e-10  # Parameter used for the Loimaranta efficiency
```

#############
# Functions #
#############

```{r}
poisson_probabilities <- function(lambda, states = default_states) {
  result <- c()
  
  for (k in 0:(states - 1)) {
    result <- c(result, dpois(k, lambda))
  }
  
  return(result)
}
```

```{r}
mat_row <- function(pos, lambda, states = default_states, steps = default_steps) {
  prob <- poisson_probabilities(lambda)
  cum_prob <- cumsum(prob)
  zero <- states - pos - 1
  vec <- c()
  rem <- states
  
  if (zero == -1) {
    vec <- c(vec, prob[1])
    rem <- rem - 1
  } else {
    vec <- c(vec, rep(0, zero), prob[1])
    rem <- rem - zero - 1
    
    if (zero != states - 2) {
      vec <- c(vec, 0)
      rem <- rem - 1
    }
  }
  
  counter <- 1
  while (rem > steps) {
    counter <- counter + 1
    vec <- c(vec, rep(0, steps - 1), prob[counter])
    rem <- rem - steps
  }
  
  if (rem != 0) {
    vec <- c(vec, rep(0, rem - 1), 1 - cum_prob[counter])
  }
  
  return(rev(vec))
}
```

```{r}
mat <- function(lambda, states = default_states) {
  mat <- matrix(nrow = states, ncol = states)
  
  for (i in 1:states) {
    mat[i,] <- mat_row(i, lambda)
  }
  
  return(mat)
}
```

```{r}
steady_states <- function(lambda) {
  eig <- eigen(t(mat(lambda)))
  steady_states <- as.numeric(eig$vectors[, 1] / sum(eig$vectors[, 1]))
  return(steady_states)
}
```

```{r}
steady_state <- function(lambda, k) {
  results <- c()
  
  for (i in 1:length(lambda)) {
    results <- c(results, steady_states(lambda[i])[k])
  }

  return(results)
}
```

```{r}
upper <- function(k, shape = default_shape, rate = default_shape, prob = default_prob) {
  return(integrate(function(x) x * steady_state(prob * x, k) * dgamma(x, shape = shape, rate = rate), 0, Inf, subdivisions = 100L)$value)
}
```

```{r}
lower <- function(k, shape = default_shape, rate = default_shape, prob = default_prob) {
  return(integrate(function(x) steady_state(prob * x, k) * dgamma(x, shape = shape, rate = rate), 0, Inf, subdivisions = 100L)$value)
}
```

```{r}
r <- function(k) {
  return(upper(k) / lower(k))
}
```

```{r}
upper <- function(k, shape = default_gamma, rate = default_gamma, prob = default_prob) {
  return(integrate(function(x) x * steady_state(prob * x, k) * dgamma(x, shape = shape, rate = rate), 0, Inf, subdivisions = 100L)$value)
}
```

```{r}
lower <- function(k, shape = default_gamma, rate = default_gamma, prob = default_prob) {
  return(integrate(function(x) steady_state(prob * x, k) * dgamma(x, shape = shape, rate = rate), 0, Inf, subdivisions = 100L)$value)
}
```

```{r}
r <- function(k) {
  return(upper(k) / lower(k))
}
```

```{r}
premiums <- function(states = default_states) {
  result <- c()
  
  for (k in 1:states) {
    result <- c(result, r(k))
  }
  
  return(result)
}
```

```{r}
calculated_premiums <- premiums()
loimaranta_efficiency <- function(lambda, premiums = calculated_premiums, infinitesimal = default_infinitesimal) {
  steady_state_premium_old <- sum(steady_states(lambda) * premiums)
  steady_state_premium_new <- sum(steady_states(lambda + infinitesimal) * premiums)
  
  result <- (log(steady_state_premium_new) - log(steady_state_premium_old)) / (log(lambda + infinitesimal) - log(lambda))
  return(result)
}
```

################
# Calculations #
################

## Assume that the number of claims in a year is Poisson distributed with
## parameter \lambda. For your system, answer the following questions for
## \lambda = 0.05.

```{r}
lambda <- 0.05
```

# a) Calculate the transition matrix P and obtain the steady-state distribution.

```{r}
# Transition matrix
mat(lambda)

# Steady-state distribution
steady_states(lambda)
```

# b) Obtain the steady-state premium.

```{r}
calculated_premiums <- premiums()
sum(steady_states(lambda) * calculated_premiums)
```

# c) Calculate the Loimaranta efficiency. Use at least two different methods to
#    numerically calculate the Loimaranta efficiency and compare the outcomes
#    of these methods. Is the system efficient for this value of \lambda? Explain.

```{r}
loimaranta_efficiency_second_method <- function(lambda, infinitesimal = default_infinitesimal) {
  mat_old <- mat(0.05 - infinitesimal)
  mat_new <- mat(0.05 + infinitesimal)
  
  for (i in 1:10) {
    mat_old <- mat_old %*% mat_old
    mat_new <- mat_new %*% mat_new
  }
  
  premiums <- premiums()
  steady_state_old <- as.vector(mat_old[1,])
  steady_state_new <- as.vector(mat_new[1,])
  steady_state_premium_old <- sum(steady_state_old * premiums)
  steady_state_premium_new <- sum(steady_state_new * premiums)
  
  result <- (log(steady_state_premium_new) - log(steady_state_premium_old)) / (log(lambda + infinitesimal) - log(lambda - infinitesimal))
  return(result)
}
```

```{r}
# Differences are minimal, so we will not discuss this in the report
loimaranta_efficiency(lambda, calculated_premiums)
loimaranta_efficiency_second_method(lambda)
```

# d) How does the Loimaranta efficiency depend on \lambda? Make a nice graph,
#    describe what it shows you and try to find an explanation for the patterns
#    that you see. Make the graph as nice as possible.

```{r}
ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  stat_function(fun = Vectorize(loimaranta_efficiency), colour = "red") +
  theme_bw() + theme(panel.grid = element_blank(), panel.grid.major.y = element_line(size = .1, color = "black")) +
  scale_y_continuous(limits = c(0, 1.2), breaks = seq(0, 1.2, by = 0.2)) +
  xlim(0, 1) + xlab("λ") + ylab("Loimaranta efficiency")
```

# e) Compare the efficiency of this bonus malus system to the Dutch system.
#    Note that a visual comparison (based on a graph) is an excellent way to
#    answer this question. Try again to make the graphs as nice as possible
#    (using colors etc).

```{r}
dutch_mat <- function(lambda) {
  prob <- poisson_probabilities(lambda)
  cum_prob <- cumsum(prob)
  
  matrix(c(1 - cum_prob[1], prob[1], rep(0, 12),
           1 - cum_prob[1], 0, prob[1], rep(0, 11),
           1 - cum_prob[1], 0, 0, prob[1], rep(0, 10),
           1 - cum_prob[1], 0, 0, 0, prob[1], rep(0, 9),
           1 - cum_prob[2], prob[2], 0, 0, 0, prob[1], rep(0, 8),
           1 - cum_prob[2], 0, prob[2], 0, 0, 0, prob[1], rep(0, 7),
           1 - cum_prob[2], 0, 0, prob[2], 0, 0, 0, prob[1], rep(0, 6),
           1 - cum_prob[2], 0, 0, 0, prob[2], 0, 0, 0, prob[1], rep(0, 5),
           1 - cum_prob[3], prob[3], 0, 0, 0, prob[2], 0, 0, 0, prob[1], rep(0, 4),
           1 - cum_prob[3], 0, prob[3], 0, 0, 0, prob[2], 0, 0, 0, prob[1], rep(0, 3),
           1 - cum_prob[3], 0, prob[3], 0, 0, 0, prob[2], 0, 0, 0, 0, prob[1], rep(0, 2),
           1 - cum_prob[3], 0, 0, prob[3], 0, 0, 0, prob[2], 0, 0, 0, 0, prob[1], rep(0, 1),
           1 - cum_prob[3], 0, 0, prob[3], 0, 0, 0, prob[2], 0, 0, 0, 0, 0, prob[1],
           1 - cum_prob[3], 0, 0, 0, prob[3], 0, 0, 0, prob[2], 0, 0, 0, 0, prob[1]),
          nrow = 14, ncol = 14, byrow = TRUE)
}
```

```{r}
dutch_steady_states <- function(lambda) {
  eig <- eigen(t(dutch_mat(lambda)))
  steady_states <- as.numeric(eig$vectors[, 1] / sum(eig$vectors[, 1]))
  return(steady_states)
}
```

```{r}
dutch_steady_state_premium <- function(lambda) {
  steady_states <- dutch_steady_states(lambda)
  percentages <- c(120, 100, 90, 80, 70, 60, 55, 50, 45, 50, 37.5, 35, 32.5, 30)
  
  return(sum(steady_states * percentages))
}
```

```{r}
dutch_loimaranta_efficiency <- function(lambda, infinitesimal = default_infinitesimal) {
  steady_state_premium_old <- dutch_steady_state_premium(lambda)
  steady_state_premium_new <- dutch_steady_state_premium(lambda + infinitesimal)
  
  result <- (log(steady_state_premium_new) - log(steady_state_premium_old)) / (log(lambda + infinitesimal) - log(lambda))
  
  return(result)
}
```

```{r}
ggplot(data = data.frame(x = 0)) +
  stat_function(aes_(colour = "Our model"), fun = Vectorize(loimaranta_efficiency)) +
  stat_function(aes_(colour = "Dutch model"), fun = Vectorize(dutch_loimaranta_efficiency)) +
  theme_bw() + theme(panel.grid = element_blank(), panel.grid.major.y = element_line(size = 0.1, color = "black")) +
  scale_y_continuous(limits = c(0, 1.2), breaks = seq(0, 1.2, by = 0.2)) +
  xlim(0, 1) + xlab("λ") + ylab("Loimaranta efficiency") + labs(color = "Model")

ggsave("comparison.png", width = 7, height = 5)
```

##########
# Other  #
##########

```{r}
default_entry_step <- 6
```

```{r}
standardized_premiums <- function(states = default_states, entry = default_entry_step) {
  result <- c()
  
  for (k in 1:states) {
    result <- c(result, r(k))
  }
  
  return(result / result[entry])
}
```

```{r}
standardized_premiums()
```

```{r}
ggplot(data = data.frame(x = 0)) +
  stat_function(fun = dgamma, args = list(default_gamma, default_gamma / default_prob), colour = "red") +theme_bw() +
  theme(panel.grid = element_blank()) +
  scale_y_continuous(limits = c(0, 15), breaks = seq(0, 15, by = 5)) +
  xlim(0, 0.25) + xlab("x") + ylab("f(x)")

ggsave("gamma.png", width = 7, height = 5)
```
